---
title: "Personal Cheat Sheet"
output:

# This is the header of this script. You can specify the format of the output here (e.g. PDF, html etc.) and also different styles, whether you want to include a table of contents, how that should look like... The possibilities are ENDLESS!

# Example 1: A nicely formatted html-notebook  
  rmdformats::material:
    css: boxes.css

# Example 2: A PDF document
#  pdf_document:
#    toc: true
    
# You can also include options for styling, e.g. for boxes:
header-includes:
    - \usepackage[most]{tcolorbox}
    - \definecolor{light-yellow}{rgb}{1, 0.95, 0.7}
    - \newtcolorbox{myquote}{colback=light-yellow,grow to right by=-10mm,grow to left by=-10mm, boxrule=0pt,boxsep=0pt,breakable}
    - \newcommand{\yellowbox}[1]{\begin{myquote} \textbf{#1} \end{myquote}}
---

```{r global_options, include = FALSE}
# Default options for this document
knitr::opts_chunk$set(echo = T,    # include code 
                      warning = F, # print warnings
                      message = F, # print messages
                      eval = F)    # evaluate code

root_dir <- "C:/Users/fried/Dropbox/.Arbeit/2021-2022_HVW_Lab/NOW/SoftwareForAnalysingData_2021/Data" # put in your home directory

knitr::opts_knit$set(root.dir = root_dir) 
```

# Intro

This is a basis for your personal R "Cheat Sheet". In the course of the semester (and beyond that) you can add all commands and "hacks" you perceive as being beneficial for you. You can just copy-paste them from the course material. In that way you can build up your personal R repertoire. 
(And additionally, it can serve as basis for your report)
Cool, isn't it? :)


# Examples for formatting 


## Bullet points

- one
- two
- three

1. blue
2. green
3. yellow

## Text

- __bold__ 
- *italics* 

## Boxes

Here is how you can include boxes. You can customize the them in the css-file "boxes.css".

```{r, include = F, eval = F}

```

:::: {.greenbox data-latex=""}
::: {.center data-latex=""}
A centered line
:::
::: {.left data-latex=""}
A left-aligned line
:::
::::

Just move this code outside the chunk into the document.
This doesn`t work for PDFs though! 
Instead, you can use commands specified in the header (they won't be displyed in html documents though).

\yellowbox{This looks like a sticky note.}

## Linebreak & new page

`<br>` inserts a line break: <br>  
`\newpage` starts a new page: 
\newpage

# R Code

## Chunks 

This code will be displayed but not evaluated (executed)...

```{r, include= TRUE, eval = FALSE}
head(iris)
```

\... while this one will also be executed.

```{r, include= TRUE, eval = TRUE}
head(iris)
```

There are many more so-called "chunk" options (e.g. `echo`, `fig.width` etc.)

## Inline-Code

You can also include results from calculations directly in your text.

```{r, eval = T}
library(apa)
tTest <- t_test(iris$Sepal.Length[iris$Species=="setosa"],
                iris$Sepal.Length[iris$Species=="versicolor"])
```

A simple calculation: `r 30*100` <br>  
Nicely formatted results of a t-test according to the guidelines of the American Psychological Association: `r apa(tTest)`

