---
title: "Simple Tests - Pratice"
output:
  
  pdf_document:
#    toc: true  #i nclude a table of contents
---

```{r setup, include = F}
knitr::opts_chunk$set(echo = F,    # include code 
                      warning = F, # print warnings
                      message = F, # print messages
                      eval = T,    # evaluate code
                      fig.width = 4.5,
                      fig.height = 3.5)

# root_dir <- "" # set working directory
# knitr::opts_knit$set(root.dir = root_dir)   
# knitr::opts_chunk$set(root.dir = root_dir)  
```
  
For the 48hr you will receive a data set and will be basically told to "Just analyse it."
That is, you should come up with some analyses to generate interesting insides and write a report including a "story line". 

This is your chance to practice conducting tests and reporting the results in R! You may use this document as a basis and just start. For each simple tests covered in the last session, pick an R dataset (run `data()` for a list), run the test and interpret the results. Report the results using in-line code, if you like.
You may also want to think about a way to visualize the results (can be very basic).
Hint: First, always think of which is your dependent variable (what you want to explain/ predict) and which is the independent variable (what you use to explain/ group) and your decision criterion (what to conclude if the p-value is larger/smaller than a threshold like 0.05).

# Example

- Dataset: ToothGrowth
- Test: t.test
- IV: `supp` (supplement), DV: `len` (tooth length)
- Decision criterion: If p $\leq$ 0.05 $\longrightarrow$ groups differ significantly

```{r load Data, include=FALSE}

# You can run your analyses in code chunks and store the results in objects 
# that you call later in-line. You may also include calculations directly inline.

library(apa)
 
ds <- ToothGrowth

t.test_sup <- t_test(ds$len[ds$supp=="OJ"], 
                     ds$len[ds$supp=="VC"])

t.test_sup

m_OJ <- round(mean(ds$len[ds$supp=="OJ"]), digits = 2)
sd_OJ <- round(sd(ds$len[ds$supp=="OJ"]), digits = 2)

m_VC <- round(mean(ds$len[ds$supp=="VC"]), digits = 2)
sd_VC <- round(sd(ds$len[ds$supp=="VC"]), digits = 2)

m_OJ; sd_OJ; m_VC; sd_VC
```

There was no significant difference in mean tooth length of guinea pigs between the group that was given orange juice (*M* = `r m_OJ`, *SD* = `r sd_OJ`) and the group that was given vitamin C (*M* = `r m_VC`, *SD* = `r sd_VC`) as supplement, `r apa(t.test_sup)`.

# Shapiro-Wilk

```{r}
# suggestion for dataset: women

ds <- women
attach(ds)
```

# F-test

```{r}
# suggestion for dataset: ToothGrowth

ds <- ToothGrowth
attach(ds)
```

# T-test

```{r}
# suggestion for dataset: sleep

ds <- sleep
attach(ds)
```

# Chi-Square

```{r}
# suggestion for dataset: infert
ds <- infert
attach(ds)
```

